---
title: "Guide to using MR-TRYX"
author: "Yoonsu Cho"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Guide to using MR-TRYX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

The following analyses should run within a couple of minutes, depending on internet speed and the traffic that the [IEU GWAS database](https://gwas.mrcieu.ac.uk/) servers are experiencing.

Begin by choosing an exposure-outcome hypothesis to explore. e.g. LDL cholesterol on coronary heart disease. These data can be extracted from the IEU GWAS database using the [TwoSampleMR](https://mrcieu.github.io/TwoSampleMR/) package:


## Data setup

If necessary install TwoSampleMR:

```r
devtools::install_github("mrcieu/TwoSampleMR@ieugwasr")
```

Create a dataset for LDL and CHD

```{r}
library(TwoSampleMR)
a <- extract_instruments("ukb-b-20175")
b <- extract_outcome_data(a$SNP, "ieu-a-7", access_token=NULL)
dat <- harmonise_data(a,b)
```

## Running Tryx

We can now perform the analysis:

This will do the following:

1. Initialise the Tryx class with the `dat` object we already created
2. Find outlier SNPs in the exposure-outcome analysis
3. Find traits in the IEU GWAS database that those outliers associate with. These traits are known as 'candidate traits'
4. Extract instruments for those 'candidate traits'
5. Perform MR of each of those 'candidate traits' against the exposure and the outcome

```{r}
library(tryx)
x <- Tryx$new(dat)
x$mrtryx()
```



### Step-by-step analysis

The analysis performed by `mrtryx()` can be broken down into multiple steps, running the following commands:

```r
# Find outlier SNPs in the exposure-outcome analysis
x$get_outliers()

# Find traits in the MR-Base database that those outliers associate with. These traits are known as 'candidate traits'
x$set_candidate_traits()
x$scan()
```

Note that you can browse available traits here: https://gwas.mrcieu.ac.uk/ and get a complete list using:

```r
traits <- TwoSampleMR::available_outcomes()
```

```r
# Extract instruments for those 'candidate traits'
x$extractions() #which includes the following functions:
  x$candidate_instruments()
  x$outcome_instruments()
  x$exposure_instruments()
  x$exposure_candidate_instruments()

# Make datasets for MR analysis
x$harmonise() #which includes the following functions:
  x$candidate_outcome_dat()
  x$candidate_exposure_dat()
  x$exposure_candidate_dat()

# Perform MR of each of those 'candidate traits' against the exposure and the outcomes
x$mr()
```


See the `?Tryx` for options on the parameters for this analysis. e.g. You can specify your own set of outliers, for example SNPs that have extreme p-values in the outcome GWAS

```r
a <- as.character(subset(dat, pval.outcome < 5e-8)$SNP)
x <- Tryx$new(dat)
x$mrtryx(outliers=a)
```

The next steps are to determine which of the candidate traits are of interest (e.g. using p-value thresholds), visualise the results, and adjust the exposure-outcome estimates based on knowledge of the 'candidate trait' associations.


## Significant candidate traits

One can determine which of the putative associations might be 'interesting' in different ways. We have provided a simple convenience function to apply different multiple testing corrections. e.g.

```{r}
x$tryx.sig()
```

Will by default use FDR of 5%. See `?Tryx` for more options.



## Adjustment

Finally, to adjust the SNP effects on the exposure and outcome traits given their influences on the candidate traits, we can run:

```{r}
x$analyse()
```

This will estimate the outlier effect that is due to the candidate pathways, adjust the outlier-outcome estimate, and re-perform MR of exposure on outcome with the adjusted outliers.

By default, this adjusts for the trait that has the largest impact for a particular SNP. 

There is also a multivariable adjustment method here, where for each outlier, all candidate trait effects are estimated jointly:

```r
x$analyse.mv()
```

NOTE: it's a good idea to check that there are no traits amongst the candidates that are identical or equivalent to the outcome: 

```r
id_remove = c("ebi-a-GCST005195", "ebi-a-GCST005840", "ebi-a-GCST005843", "ieu-a-1110", "ieu-a-798", "ukb-b-1061", 
"ukb-b-11590", "ukb-b-11895", "ukb-b-12014", "ukb-b-12465", "ukb-b-12493", "ukb-b-12646", "ukb-b-13352", "ukb-b-14057", 
"ukb-b-14177", "ukb-b-14371", "ukb-b-14395", "ukb-b-15169", "ukb-b-15491", "ukb-b-15686", "ukb-b-15748", "ukb-b-15829", 
"ukb-b-16606", "ukb-b-1668", "ukb-b-17360", "ukb-b-18009", "ukb-b-18018", "ukb-b-18167", "ukb-b-18408", "ukb-b-18799", 
"ukb-b-20300", "ukb-b-20379", "ukb-b-3656", "ukb-b-4063", "ukb-b-7137", "ukb-b-7436", "ukb-b-7869", "ukb-b-7992", 
"ukb-b-8468", "ukb-b-8650", "ukb-b-8746", "ukb-b-8755", "ukb-b-9207", "ukb-b-10454", "ukb-b-11632", "ukb-b-11740", 
"ukb-b-11971", "ukb-b-16866", "ukb-b-19456", "ukb-b-8778", "ukb-b-12019")

x$analyse.mv(id_remove = id_remove)
```

The adjusted effect estimates:

```{r}
x$output$analysis$estimates
```



## Visualisation


### Adjustment plot

A plot is generated showing how SNP effects have changed due to candidate trait adjustments in:


```{r}
x$output$analysis$plot
```


### Network plot

To produce a basic diagram of the connectivity of SNPs, candidate traits, exposure and outcome:

```r
tryx.network(x$output)
```

This shows that some candidate traits influence the exposure only, the outcome only, or both the exposure and the outcome. 

### Volcano plot

You can also create a volcano plot of the candidate-exposure and/or candidate-outcome associations. e.g. to show exposures and outcomes

```{r}
volcano_plot(
    rbind(x$output$candidate_exposure_mr, x$output$candidate_outcome_mr)
)
```

or e.g. just exposures

```{r}
volcano_plot(x$output$candidate_exposure_mr)
```


### Manhattan plot

Similar to the volcano plot, a Manhattan plot can also be produced. This can be done for the candidate-exposure associations

```{r}
x$manhattan_plot("exposure")
x$output$plots$manhattan_plot
```

or the candidate-outcome associations

```{r}
x$manhattan_plot("outcome")
x$output$plots$manhattan_plot
```
